---
title: "PaperMicroUltimate"
output: html_document
date: "2024-10-04"
---


```{r echo=FALSE, warning=FALSE}
suppressMessages(library(RColorBrewer))
suppressMessages(library(ggExtra))
suppressMessages(library(openxlsx))
suppressMessages(library(dplyr))
suppressMessages(library(pheatmap))
suppressMessages(library(ggplot2))
suppressMessages(library(ggrepel))
suppressMessages(library(gridExtra))
suppressMessages(library(patchwork))
suppressMessages(library(reshape2))
suppressMessages(library(colorRamp2))
suppressMessages(library(viridis))
suppressMessages(library(edgeR))
suppressMessages(library(readr))
suppressMessages(library(vegan))
suppressMessages(library(dendextend))
suppressMessages(library(tidyr))
suppressMessages(library(sva))
suppressMessages(library(rstudioapi))
suppressMessages(library(eulerr))
```

```{r echo=FALSE, warning=FALSE}
script_dir <- dirname(getActiveDocumentContext()$path)
setwd(script_dir)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = script_dir)
```
In this section, we import and preprocess the key dataframes required for the analysis of the **vulvar microbiome**. The datasets include microbiome absolute abundances generated using Kraken, samples metadata, and gene expression counts, all of which are crucial for subsequent statistical and bioinformatic analysis.

-   **Kraken Output**: The dataframe `kdata` represents the taxonomic abundances from the Kraken output. To eliminate potential human DNA contamination, reads associated with *Homo sapiens* have been filtered out. Additionally, a pseudo-count of 1 has been added to all entries to avoid zeroes during log transformation or other mathematical operations.

-   **Metadata**: The `meta` dataframe includes relevant clinical and experimental information about the cohort, containing details for each sample in the study, such as patient/control status.



```{r echo=FALSE}
#Dataframe metadata
meta=read.delim("../local/data/df_metadataPRJEB61325.txt")
meta$site_adj=meta$SwabSite
meta[grepl("labia majora",meta$site_adj,ignore.case = T),]$site_adj="labia majora"
meta[grepl("labia minora",meta$site_adj,ignore.case = T),]$site_adj="labia minora"

#Dataframe output Kraken
kdata=read.delim("../local/data/bracken_merged_abbundances.num_PRJEB61325.txt", sep= "\t")
kdata = kdata[!grepl("sapiens",rownames(kdata)),]
rownames(kdata)=kdata$name
kdata= kdata[,-c(1:3)]
kdata=kdata+1
kdata=kdata %>% dplyr::select(-matches("\\.1$"))
idx = which(rowSums(kdata >= 50) >= ncol(kdata)/4)
kdata_filt= as.data.frame(kdata[idx,])


kdataus=read.delim("../results/countsToT_decont.txt",sep= "\t")
kdataus=kdataus+1
samp=read.xlsx("../local/data/Vulvodinia_Samples.xlsx")


kdataH=kdataus[,grepl("H_",colnames(kdataus))]

idxH= which(rowSums(kdataH >= 50) >= ncol(kdataH)/4)
kdataus_filt= as.data.frame(kdataH[idxH,])


final <- merge(kdataus,kdata,by=0)
final=final[,!grepl("D_",colnames(final))]
final <- final[,!grepl("TT81",colnames(final))] 
final <- final[,!grepl("TT60",colnames(final))]
final <- final[,!grepl("H_BP3",colnames(final))]
rownames(final)=final$Row.names
final$Row.names=NULL
```

```{r}
# ##NORM
final_frac<-sweep(final,2,colSums(final),"/")
rvfinal <- as.data.frame(t(final_frac))
```

```{r}

set.seed(123)  # For reproducibility

NMDS <- metaMDS(rvfinal, distance = "bray", trymax = 100, autotransform = T, k = 4)

# stress value around or above 0.2 is deemed suspect
# stress values equal to or below 0.1 are considered fair
# stress values equal to or below 0.05 indicate good fit

score <- as.data.frame(scores(NMDS)$site)
score$sampleID <- rownames(score)
score <- merge(score, meta, by.x  = "sampleID",by.y = "run_accession" )
score = filter(score,!(INSDC.status %in% c("NC1", "NC2", "NC3", "NC4","PC1","PC2","PC3","PC4")))
colors <- brewer.pal(n = length(unique(score$SwabSite)), name = "Set1")
Vulv_H<- score %>% filter(INSDC.status =="Healthy") %>% pull("sampleID")
All_D <- score %>% filter(site == "vaginal" | site =="anal") %>% pull("sampleID")  
VulvaUS<- score %>% filter(site =="vulvar_us") %>% pull("sampleID")
Anal <- score %>% filter(site =="anal") %>% pull("sampleID")  
Vaginal <- score %>% filter(site == "vaginal" ) %>% pull("sampleID")  
idx<- which(score$sampleID %in% unique(c(Vulv_H,All_D)))

color1 <- brewer.pal(n = length(unique(score[idx,]$site_adj)), name = "Set1")
score1=as.data.frame(score[idx,])

ggplot() +
  geom_jitter(data = score1, aes(x = NMDS1, y = NMDS2,shape = batch, color = site), 
             size = 3, stroke = 0.5, alpha = 0.9) +  # Transparency and stroke for outline control
  scale_color_manual(values = colors) +  # Softer color palette for the fill
  labs(title = "Plot Lichen Study",
       x = "NMDS2", 
       y = "NMDS3",
       fill = "Site", 
       shape = "Status") +  # Add axis and legend labels
  theme_minimal(base_size = 14)  # Clean theme

```

```{r}


mod<-model.matrix(~batch, meta[match(colnames(final_frac),meta$run_accession),])
combdata<-as.data.frame(ComBat(dat=final_frac,batch = meta[match(colnames(final_frac),meta$run_accession),]$batch),mod=mod)


NMDS <- metaMDS(t(combdata), distance = "bray", trymax = 100, autotransform = T, k = 4)

# stress value around or above 0.2 is deemed suspect
# stress values equal to or below 0.1 are considered fair
# stress values equal to or below 0.05 indicate good fit

# library(tidyverse)
score <- as.data.frame(scores(NMDS))
score$sampleID <- rownames(score)

score <- merge(score, meta, by.x  = "sampleID",by.y = "run_accession" )
score = filter(score,!(INSDC.status %in% c("NC1", "NC2", "NC3", "NC4","PC1","PC2","PC3","PC4")))




Vulv_H<- score %>% filter(INSDC.status =="Healthy") %>% pull("sampleID")
All_D <- score %>% filter(site == "vaginal" | site =="anal") %>% pull("sampleID")  
VulvaUS<- score %>% filter(site =="vulvar_us") %>% pull("sampleID")
Anal <- score %>% filter(site =="anal") %>% pull("sampleID")  
Vaginal <- score %>% filter(site == "vaginal" ) %>% pull("sampleID")  
idx<- which(score$sampleID %in% unique(c(Vulv_H,All_D)))

color1 <- brewer.pal(n = length(unique(score[idx,]$site_adj)), name = "Set1")
score1=as.data.frame(score[idx,])
colors <- c("#B6DC76","#50808E","#D88373","#D66")
ggplot()+
  geom_jitter(data = score1, aes(x = NMDS1, y = NMDS2,shape = batch, color = site_adj), 
             size = 5, stroke = 0.5, alpha = 0.9) +  # Transparency and stroke for outline control
  scale_color_manual(values = color1) +  # Softer color palette for the fill
  # scale_shape_manual(values = c(21, 24, 23, 22)) +  # Hollow shapes with fillable areas
  labs(title = "Comparison with other shotgun-metagenome analysis",
       x = "NMDS1", 
       y = "NMDS2",
       color = "Site", 
       shape = "Studies") +  # Add axis and legend labels
  theme_minimal(base_size = 14)  # Clean theme

p=ggplot() +
  geom_jitter(data = score1, aes(x = NMDS1, y = NMDS2, shape = batch, color = site), 
              size = 7, alpha = 1) +
  stat_ellipse(data = score1, aes(x = NMDS1, y = NMDS2, group = site, fill = site), 
               type = "norm", level = 0.96, geom = "polygon", alpha = 0.2) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) + 
  scale_shape_manual(values = c(16,18))+
  labs(title = "Comparison with other shotgun-metagenome analysis",
       x = "NMDS1", 
       y = "NMDS2",
       color = "Site", 
       shape = "Studies") +
  theme_minimal(base_size = 14)
 ggMarginal(p, type = "density", groupColour = TRUE, groupFill = TRUE)


```

```{r}
select_top_100 <- function(df) {
  df %>%
    mutate(TotalAbundance = rowSums(df, na.rm = TRUE)) %>%  # Calcola la somma delle abbondanze per ogni riga
    arrange(desc(TotalAbundance)) %>%  # Ordina in ordine decrescente
    slice_head(n = 100) %>%  # Seleziona le prime 100 righe
    dplyr::select(-TotalAbundance)  # Rimuovi la colonna di supporto
}

VaginalDB<-combdata[,Vaginal]
AnalDB<-combdata[,Anal]
VulvarDB<-combdata[,VulvaUS]

# Applica la funzione ai tre dataset
VaginalDB_top100 <- select_top_100(VaginalDB)
AnalDB_top100 <- select_top_100(AnalDB)
VulvarDB_top100 <- select_top_100(VulvarDB)

VaginalDB_top100_means <- rowMeans(VaginalDB_top100)
AnalDB_top100_means <- rowMeans(AnalDB_top100)
VulvarDB_top100_means <- rowMeans(VulvarDB_top100)
first <- merge(VulvarDB_top100_means,AnalDB_top100_means,by=0,all=T) %>% tibble::column_to_rownames("Row.names") %>% dplyr::rename(Vulvar=x,Anal=y)
all <- merge(first,VaginalDB_top100_means,by=0,all=T)%>% tibble::column_to_rownames("Row.names") %>% dplyr::rename(Vaginal=y) 
all[is.na(all)] <- 0
write.table(all,"../results/mergeanvulvag.txt",sep="\t")

venn_data <- list(
  Vaginal = rownames(VaginalDB_top100),
  Anal = rownames(AnalDB_top100),
  Vulvar = rownames(VulvarDB_top100)
)

# Calcola il diagramma Euler proporzionale
fit <- euler(venn_data)
# Crea il plot con stile avanzato
plot(fit,
     quantities = list(type = "counts", font = 2, cex = 1.2, col = "gray30"),
     fills = list(fill = c("#B6DC76","#50808E","#D66"), alpha = 0.7),
     edges = list(col = "gray30", lex = 2),
     main = "Overlap Top 100 Abundance",
     legend = list(side = "right"))
```

```{r alphadiversity}
nreads_min<-min(colSums(final))
nreads_max<-max(colSums(final))
fold<-round(nreads_max/nreads_min,0)

kdata_subs<-rrarefy(t(final),nreads_min)
alpha_subs<-diversity(kdata_subs)
alphasubs=data.frame(samples=names(alpha_subs),value=alpha_subs)

ggplot(alphasubs, aes(x = samples, y = factor(1), size = value, colour = value)) +
  geom_point() +
  scale_size_continuous(range = c(0, 8)) +
  scale_color_gradient(low = "grey90", high = "#D94943") +  # Intensity of color
  theme_bw() +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 5, angle = 45),  # Riduce la dimensione e inclina le etichette dell'asse y
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none"  # Hide the legend
  )+
  coord_fixed(ratio = 2)
```

```{r}
association <- read.csv("../local/data/MicroPhenoDBAssociationScores.csv") %>% select(-Pmid, -Association_ID)
key_cols <- setdiff(names(association), "Score")


merged_dupl <- merge(final_frac, association, by.x = 0, by.y = "Microbe")

# Compute the mean of 'Score' while grouping by 'Row.names'
Scores <- merged_dupl %>%
    group_by(Row.names) %>%
    summarise(Score = mean(Score, na.rm = TRUE), .groups = "drop") %>%
    tibble::column_to_rownames("Row.names")

merged <- merge(final_frac,Scores,by=0)%>%tibble::column_to_rownames("Row.names")


diseasescore<-list()
for (names in setdiff(colnames(merged), c("Score","Genus"))) {

diseasescore[names]<-sum(merged[,names]*merged[,"Score"])  
  
}

disbiodb=data.frame(samples=names(unlist(diseasescore)),dysbio=unlist(diseasescore))
disbiodb$samples=factor(disbiodb$samples)
disbiodb$dysbio<-disbiodb$dysbio

ggplot(disbiodb, aes(x = samples, y = factor(1), size = dysbio, colour = dysbio)) +
  geom_point() +
  scale_size_continuous(range = c(0, 5)) +
  scale_color_gradient(low = "grey90", high = "#0070BB") +  # Intensità del colore
  theme_bw() +
  theme(
    axis.title.y = element_blank(),
    axis.text.y =  element_blank(),
    axis.text.x = element_text(size = 5, angle = 45), 
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none"  # Nasconde la legenda
  ) +
  coord_fixed(ratio = 2)

```

