---
title: 'Project: vulvarmicrob'
author: 'Emanuele Monteleone'
date: '`r Sys.Date()`'
output: html_document
---



Setup your environment...

```{r}
# Environment variables for R

PROJECT_DIR="vulvarmicrob"
DATA_DIR=file.path(PROJECT_DIR,"local/data")
SRC_DIR=file.path(PROJECT_DIR,"local/src")
RES_DIR=file.path(PROJECT_DIR,"local/results")

## End Setup

```

library used in this analysis

```{r}
suppressPackageStartupMessages(library(WGCNA))
suppressPackageStartupMessages(library(openxlsx))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tdyr))
suppressPackageStartupMessages(library(tibble))
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(enrichR))
suppressPackageStartupMessages(library(circlize))
suppressPackageStartupMessages(library(purrr))
suppressPackageStartupMessages(library(vegan))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(biomaRt))
suppressPackageStartupMessages(library(org.Hs.eg.db))

```

Helper Functions

```{r}
#Edge List Intra-Domain (Host & Microbe)
create_intra_domain_edges <- function(cor_results, source_col_name, target_col_name, interaction_type) {
  
  cor_matrix <- cor_results$cor
  p_matrix <- cor_results$p
  
  edges_long <- cor_matrix %>%
    as.data.frame() %>%
    tibble::rownames_to_column(source_col_name) %>%
    tidyr::pivot_longer(cols = -source_col_name, 
                        names_to = target_col_name, 
                        values_to = "Correlation_R")
  
  p_long <- p_matrix %>%
    as.data.frame() %>%
    tibble::rownames_to_column(source_col_name) %>%
    tidyr::pivot_longer(cols = -source_col_name, 
                        names_to = target_col_name, 
                        values_to = "P_Value")
  
  edge_table_raw <- inner_join(edges_long, p_long, 
                               by = c(source_col_name, target_col_name))
  
  edge_table_filtered <- edge_table_raw %>%
    dplyr::filter(P_Value < P_THRESHOLD, 
                  abs(Correlation_R) >= R_THRESHOLD,
                  !!sym(source_col_name) != !!sym(target_col_name)) %>% # Removes A <-> A
    
    # Removes duplicates (keep only  A <-> B and not B <-> A)
    dplyr::rowwise() %>%
    dplyr::filter(!!sym(source_col_name) < !!sym(target_col_name)) %>% 
    dplyr::ungroup() %>%
    
    dplyr::mutate(
      Source = !!sym(source_col_name),
      Target = !!sym(target_col_name),
      Interaction = interaction_type,
      Direction = ifelse(Correlation_R > 0, "Positiva", "Negativa")
    ) %>%
    dplyr::select(Source, Target, Interaction, Correlation_R, P_Value, Direction)
  
  return(edge_table_filtered)
}

#Edge List Inter-Domain (Host ↔ Microbe)
create_inter_domain_edges <- function(cor_results, interaction_type) {
  
  cor_matrix <- cor_results$cor
  p_matrix <- cor_results$p
  
  cor_long <- cor_matrix %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Host_Gene") %>%
    tidyr::pivot_longer(cols = -Host_Gene, 
                        names_to = "Microbe_KO", 
                        values_to = "Correlation_R")
  
  p_long <- p_matrix %>%
    as.data.frame() %>%
    tibble::rownames_to_column("Host_Gene") %>%
    tidyr::pivot_longer(cols = -Host_Gene, 
                        names_to = "Microbe_KO", 
                        values_to = "P_Value")
  
  edge_table_raw <- inner_join(cor_long, p_long, 
                               by = c("Host_Gene", "Microbe_KO"))
  
  edge_table_filtered <- edge_table_raw %>%
    dplyr::filter(P_Value < P_THRESHOLD, 
                  abs(Correlation_R) >= R_THRESHOLD) %>%
    
    dplyr::mutate(
      Source = Host_Gene,
      Target = Microbe_KO,
      Interaction = interaction_type,
      Direction = ifelse(Correlation_R > 0, "Positiva", "Negativa")
    ) %>%
    dplyr::select(Source, Target, Interaction, Correlation_R, P_Value, Direction)
  
  return(edge_table_filtered)
}

assign_function <- function(gene) {
  if (gene %in% immune) return("Immune")
  if (gene %in% kerat) return("Keratinization")
  if (gene %in% apo) return("apoptotic process")
  if (gene %in% inate_immune) return("innate immune response")
  if (gene %in% cell_migration) return("cell migration")
  if (gene %in%cell_diff) return("epithelial cell differentiation")
  if (gene %in% wound) return("wound healing")
  if (gene %in% reg_infl) return("regulation of inflammatory response")
  return("Other/General")
}

getGene<-function(anno_file, term) {
  gene_list <- anno_file |> dplyr::filter(PROCESS_NAME==term) |> 
  pull(SYMBOL) |> unique()
  return(gene_list)
}
```


load count data and metadata

```{r}
counts<-read.xlsx(file.path(DATA_DIR,"GEP.count.xlsx"),rowNames = T)

meta<-read.xlsx(file.path(DATA_DIR,"meta.xlsx"),rowNames = T)
counts=counts[,match(rownames(meta),colnames(counts))]
```


Normalize and filter hosta gene expression data using DESeq2 
```{r}
dds_rnaseq<-DESeqDataSetFromMatrix(countData = as.matrix(counts),colData = meta, design = ~1)
nreads <- 10
nsamples <- 10
keep <- rowSums(counts(dds_rnaseq) >= nreads) >= nsamples
dds_rnaseq <- dds_rnaseq[keep,]
dds_rnaseq<-estimateSizeFactors(dds_rnaseq)
```

Normalized microbiome data using CLR normalization
```{r}
counts_braken<-read.delim(file.path(DATA_DIR,"bracken_merged_abundances.num.filtered.nocontam.txt"))
colnames(counts_braken)<-gsub("H_","",colnames(counts_braken))
counts_braken_comp <- sweep(counts_braken, 2, colSums(counts_braken), "/")
clr_microbiome_data <- decostand(counts_braken_comp+0.0001, method="clr")
gene_matrix=t(counts(dds_rnaseq,normalized=T))
microbe_matrix=t(t(clr_microbiome_data))
```


Calculate correlation between gene expression and microbiome composition using spearman correlation

```{r}
cor_rnaseq_microbiome=corAndPvalue(gene_matrix,microbe_matrix,method = "spearman")
results=reshape2::melt(cor_rnaseq_microbiome$cor)
results$pval=reshape2::melt(cor_rnaseq_microbiome$p)$value
colnames(results)<-c("Gene","microorg","cor","cor_pval")
significant_correlations <- results[which(results$cor_pval < 0.05 & abs(results$cor) > 0.6), ]
```

Calculate correlation between microbes and separately between host genes

```{r}
# clr + spearman
cor_microbiome <- corAndPvalue(t(clr_microbiome_data),method = "spearman")
cor_host<-corAndPvalue(gene_matrix,method = "spearman")
```


Filter and normalize the data and perform WGCNA analysis

```{r}
# filt and normalize again using varianceStabilizingTransformation (DESeq2)
dds_rnaseq_vst<-varianceStabilizingTransformation(dds_rnaseq[rownames(dds_rnaseq)%in%significant_correlations$Gene,])
enableWGCNAThreads(nThreads =7)
powers = c(c(1:10), seq(from = 12, to = 20, by = 2))
sft = pickSoftThreshold(t(assay(dds_rnaseq_vst)), powerVector = powers,verbose=5)
par(mfrow = c(1,2));
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, col = "red"
)
abline(h = 0.90, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers, col = "red")
```


Based on the scale independence and mean connectivity, we choose a power of 7

```{r}
picked_power=7
```


Perform WGCNA analysis

```{r}
netwk <- blockwiseModules(t(assay(dds_rnaseq_vst)),                
                          # == Adjacency Function ==
                          power = picked_power,
                          networkType = "signed",
                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = 30,
                          maxBlockSize = 4000,
                          # == Module Adjustments ==
                          reassignThreshold = 0,
                          mergeCutHeight = 0.25,
                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = T,
                          saveTOMFileBase = "ER",
                          # == Output Options
                          numericLabels = T,
                          verbose = 3)
# Convert labels to colors for plotting
mergedColors = labels2colors(netwk$colors)

```


Plot the module-trait relationships; traits are clusters based on the microbiome composition

```{r}
# Get Module Eigengenes per cluster
MEs0 <- moduleEigengenes(t(assay(dds_rnaseq_vst)), mergedColors)$eigengenes
# Reorder modules so similar modules are next to each other
MEs0 <- orderMEs(MEs0)
module_order = names(MEs0) %>% gsub("ME","", .)
# Add treatment names
cl_micro<-read.delim(file.path(DATA_DIR,"cluster_healthy_microbiome.txt"),sep=";",row.names = 1)
rownames(cl_micro)=gsub("H_","",rownames(cl_micro))
MEs0 = merge(cl_micro,MEs0,by=0)
MEs0 = MEs0 %>% tibble::column_to_rownames(var="Row.names")

design=model.matrix(~0+MEs0$Dominance)
moduleTraitCor = cor(MEs0[,-c(1,2)], design , use = "p");
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nrow(t(assay(dds_rnaseq_vst))))
textMatrix = paste(signif(moduleTraitCor, 2), "\n(",
                   signif(moduleTraitPvalue, 1), ")", sep = "");
dim(textMatrix) = dim(moduleTraitCor)
par(mar = c(6, 8, 3, 3));
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = c("Lactobacillus crispatus","Lactobacillus gasseri",
                           "Lactobacillus iners","Others"),
               yLabels = names(MEs0[,-c(1,2)]),
               ySymbols = gsub("ME","",names(MEs0[,-c(1,2)])),
               colorLabels = FALSE,
               colors = blueWhiteRed(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(-1,1),
               main = paste("Module-trait relationships"))
dev.off()

```


Perform EnrichR analysis for each module

```{r}
dbs<-c("GO_Biological_Process_2025","Elsevier_Pathway_Collection","Reactome_Pathways_2024")
websiteLive <- getOption("enrichR.live")
module_df <- data.frame(
  gene_id = names(netwk$colors),
  colors = labels2colors(netwk$colors)
)
#split by color
module_df_list<-split(module_df$gene_id,module_df$colors)
if (websiteLive) {
  enriched_degs <- lapply(module_df_list, function(x) enrichr(x, dbs))
}
```

Create network for Cytoscape analysis

```{r}
P_THRESHOLD <- 0.05
R_THRESHOLD <- 0


#  Host-Host 
edges_host <- create_intra_domain_edges(
  cor_host, 
  source_col_name = "Host_Gene", 
  target_col_name = "Target_Gene", 
  interaction_type = "Host-Host Cor"
)

# Microbe-Microbe Correlation
edges_microbe <- create_intra_domain_edges(
  cor_microbiome, 
  source_col_name = "Source_KO", 
  target_col_name = "Target_KO", 
  interaction_type = "Microbe-Microbe Cor"
)

#  Host-Microbe Correlation
edges_host_microbe <- create_inter_domain_edges(
  cor_microbiome_host, 
  interaction_type = "Host-Microbe Cor"
)

final_network_edges <- bind_rows(edges_host, edges_microbe, edges_host_microbe)
node_table<-rbind(data.frame(Source=rownames(cor_host$cor), Type="Gene"),
                  data.frame(Source=rownames(cor_microbiome$cor), Type="Microbe")
)
# Write the node list to a CSV file

write.csv(node_table,
          file.path(PROJECT_DIR, "results/node_list.csv"),
          row.names = FALSE, quote = FALSE)
# Write the edge list to a CSV file

write.csv(final_network_edges, 
          file.path(PROJECT_DIR, "results/edge_list.csv"), 
          row.names = FALSE, quote = FALSE)


```

Microbial functional impact analysis

``` {r}
hm_edges <- final_network_edges %>%
  dplyr::filter(Interaction == "Host-Microbe Cor")
#get GO BP annotations
ann_go <- select(org.Hs.eg.db,
     keys = unique(hm_edges$Source),
     columns = c("SYMBOL", "GO"),
     keytype = "SYMBOL")
ann_go_bp <- subset(ann_go, ONTOLOGY == "BP")
ann_go_bp$PROCESS_NAME <- Term(ann_go_bp$GO)
#get annotation for key biological processes in the epithelium

kerat<-getGene(ann_go_bp,"keratinization")
immune<-unique(c(getGene(ann_go_bp,"immune system process"),
                 getGene(ann_go_bp,"regulation of immune system process")
))
cell_diff<-getGene(ann_go_bp,"epithelial cell differentiation")
chr_remod<-getGene(ann_go_bp,"chromatin remodeling")
wound<-getGene(ann_go_bp,"wound healing")
reg_infl<-getGene(ann_go_bp,"regulation of inflammatory response")
apo<-c(getGene(ann_go_bp,"apoptotic process"),getGene(ann_go_bp,"epithelial cell apoptotic process"))
inate_immune<-getGene(ann_go_bp,"innate immune response")
cell_migration<-getGene(ann_go_bp,"cell migration")
cell_surface<-getGene(ann_go_bp,"cell surface receptor signaling pathway")
#assign function to genes
hm_edges$Host_Function <- sapply(hm_edges$Source, assign_function)

# top 10% microbes based on average composition across all samples
microbes<-c(
"Cutibacterium acnes"
"Fannyhessea vaginae"
"Gardnerella swidsinkii"
"Gardnerella vaginalis"
"Lactobacillus crispatus"
"Lactobacillus gasseri"
"Lactobacillus iners"
"Lactobacillus jensenii"
"Prevotella bivia")

plot_data_raw <- hm_edges %>%
  dplyr::filter(Host_Function != "Other/General", Target %in% microbes, abs(Correlation_R)>0.5) %>%
  dplyr::filter(!grepl("chromatin", Host_Function)) |>
  group_by(Target, Host_Function, Direction) %>%
  summarise(count = n(), .groups = 'drop')
total_per_microbe <- plot_data_raw %>%
  group_by(Target) %>%
  summarise(total_n = sum(count))
functional_direction_data <- plot_data_raw %>%
  left_join(total_per_microbe, by = "Target") %>%
  mutate(percentage = (count / total_n) * 100) %>%
  mutate(percentage_plot = ifelse(Direction == "Negativa", -percentage, percentage)) %>%
  mutate()

#Diverging Barplot 
ggplot(functional_direction_data, aes(x = reorder(Target, percentage_plot, FUN = sum), 
                                      y = percentage_plot, 
                                      fill = Host_Function)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  coord_flip() +
  scale_fill_brewer(palette = "Set3") +
  scale_y_continuous(labels = function(x) paste0(abs(x), "%"), 
                     limits = c(-100, 100)) +
  labs(title = "Microbial Functional Impact (GO terms)",
       subtitle = "Negative Correlations (Left) vs Positive Correlations (Right)",
       x = "Microorganism",
       y = "Percentage of Interactions (%)",
       fill = "Biological Process (GO)") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8))
```

Bridges analysis: host nodes exhibiting statistically significant interactions with two or more distinct microbial taxa
```{r}
bridges <- hm_edges %>%
  dplyr::filter(Host_Function != "Other/General", Target %in% microbes$V1) %>%
  group_by(Source) %>%
  summarise(
    n_microbes = n_distinct(Target),
    microbe_list = paste(unique(Target), collapse = ", "),
    functions = unique(Host_Function),
    total_interactions = n()
  ) %>%
  dplyr::filter(n_microbes > 1) %>% 
  arrange(desc(n_microbes))
bridge_genes <- hm_edges %>%
  dplyr::filter(Host_Function != "Other/General", Target %in% microbes) %>%
  group_by(Source) %>%
  mutate(n_microbes = n_distinct(Target)) %>%
  dplyr::filter(n_microbes > 1) %>%
  ungroup()
bridge_genes_ordered <- bridge_genes %>%
  arrange(Host_Function)
row_ann_ordered <- data.frame(
  unique(data.frame(Source=bridge_genes_ordered$Source,Host_Function=bridge_genes_ordered$Host_Function))
)
rownames(row_ann_ordered)<-row_ann_ordered$Source
row_ann_ordered$Source<-NULL
gaps <- cumsum(table(row_ann_ordered$Host_Function)[unique(row_ann_ordered$Host_Function)])
matrix_data <- bridge_genes %>%
  dplyr::select(Source, Target, Correlation_R) %>%
  pivot_wider(names_from = Target, values_from = Correlation_R, values_fill = 0) %>%
  column_to_rownames("Source")
matrix_ordered <- matrix_data[unique(bridge_genes_ordered$Source), ]

heatmap_colors <- colorRampPalette(c("navy", "white", "firebrick3"))(100)

pheatmap(t(matrix_ordered),
         annotation_col = row_ann_ordered,
         color = heatmap_colors,
         breaks = seq(-1, 1, length.out = 101),
         main = "Gene Bridges: Host-Microbe Co-regulation",
         clustering_distance_cols = "euclidean",
         clustering_method="ward.D2",
          
         show_colnames = TRUE,
         gaps_col  = gaps,
         cluster_cols = F,
         fontsize_row = 8,
         fontsize_col = 10,
         border_color = NA)
```



