---
title: 'Project: vulvarmicrob'
author: 'Emanuele Monteleone'
date: '`r Sys.Date()`'
output: html_document
---



Setup your environment...

```{r}
# Environment variables for R

PROJECT_DIR="vulvarmicrob"
DATA_DIR=file.path(PROJECT_DIR,"local/data")
SRC_DIR=file.path(PROJECT_DIR,"local/src")
RES_DIR=file.path(PROJECT_DIR,"local/results")

## End Setup

```

library used in this analysis

```{r}
suppressPackageStartupMessages(library(WGCNA))
suppressPackageStartupMessages(library(openxlsx))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(DESeq2))
suppressPackageStartupMessages(library(enrichR))
suppressPackageStartupMessages(library(circlize))
suppressPackageStartupMessages(library(purrr))
```


load count data and metadata

```{r}
counts<-read.xlsx(file.path(DATA_DIR,"GEP.count.xlsx"),rowNames = T)

meta<-read.xlsx(file.path(DATA_DIR,"meta.xlsx"),rowNames = T)
counts=counts[,match(rownames(meta),colnames(counts))]
```


Normalize and filter data using DESeq2 for both host and microbiome RNASeq data

```{r}
dds_rnaseq<-DESeqDataSetFromMatrix(countData = as.matrix(counts),colData = meta, design = ~1)
nreads <- 10
nsamples <- 10
keep <- rowSums(counts(dds_rnaseq) >= nreads) >= nsamples
dds_rnaseq <- dds_rnaseq[keep,]
dds_rnaseq<-estimateSizeFactors(dds_rnaseq)
#microbiome
braken<-read.delim(file.path(DATA_DIR,"bracken_merged_abbundances.num.filtered.txt"))
rownames(braken)=braken$name
braken$name=NULL
braken<-braken[!grepl("taxonomy",colnames(braken))]
counts_braken=braken[,match(rownames(meta),colnames(braken))]
dds_microbiome<-DESeqDataSetFromMatrix(countData = as.matrix(counts_braken),colData = meta, design = ~1)
dds_microbiome<-estimateSizeFactors(dds_microbiome)
gene_matrix=t(counts(dds_rnaseq,normalized=T))
microbe_matrix=t(counts(dds_microbiome,normalized=T))
```


Calculate correlation between gene expression and microbiome composition using spearman correlation

```{r}
cor_rnaseq_microbiome=corAndPvalue(gene_matrix,microbe_matrix,method = "spearman")
results=reshape2::melt(cor_rnaseq_microbiome$cor)
results$pval=reshape2::melt(cor_rnaseq_microbiome$p)$value
colnames(results)<-c("Gene","microorg","cor","cor_pval")
significant_correlations <- results[which(results$cor_pval < 0.05 & abs(results$cor) > 0.6), ]
```


Filter and normalize again the data and perform WGCNA analysis

```{r}
# filt and normalize again 
dds_rnaseq_vst<-varianceStabilizingTransformation(dds_rnaseq[rownames(dds_rnaseq)%in%significant_correlations$Gene,])
#similarity_matrix <- WGCNA::cor(t(counts(dds_rnaseq,normalized=T)))
enableWGCNAThreads(nThreads =7)
powers = c(c(1:10), seq(from = 12, to = 20, by = 2))
sft = pickSoftThreshold(t(assay(dds_rnaseq_vst)), powerVector = powers,verbose=5)
par(mfrow = c(1,2));
plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = paste("Scale independence")
)
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, col = "red"
)
abline(h = 0.90, col = "red")
plot(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity")
)
text(sft$fitIndices[, 1],
     sft$fitIndices[, 5],
     labels = powers, col = "red")
```


Based on the scale independence and mean connectivity, we choose a power of 7

```{r}
picked_power=7
```


Perform WGCNA analysis

```{r}
netwk <- blockwiseModules(t(assay(dds_rnaseq_vst)),                
                          # == Adjacency Function ==
                          power = picked_power,
                          networkType = "signed",
                          # == Tree and Block Options ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = 30,
                          maxBlockSize = 4000,
                          # == Module Adjustments ==
                          reassignThreshold = 0,
                          mergeCutHeight = 0.25,
                          # == TOM == Archive the run results in TOM file (saves time)
                          saveTOMs = T,
                          saveTOMFileBase = "ER",
                          # == Output Options
                          numericLabels = T,
                          verbose = 3)
# Convert labels to colors for plotting
mergedColors = labels2colors(netwk$colors)

```


Plot the module-trait relationships; traits are clusters based on the microbiome composition

```{r}
# Get Module Eigengenes per cluster
MEs0 <- moduleEigengenes(t(assay(dds_rnaseq_vst)), mergedColors)$eigengenes
# Reorder modules so similar modules are next to each other
MEs0 <- orderMEs(MEs0)
module_order = names(MEs0) %>% gsub("ME","", .)
# Add treatment names
cl_micro<-read.delim(file.path(DATA_DIR,"cluster_healthy_microbiome.txt"),sep=";",row.names = 1)
rownames(cl_micro)=gsub("H_","",rownames(cl_micro))
MEs0 = merge(cl_micro,MEs0,by=0)
MEs0 = MEs0 %>% tibble::column_to_rownames(var="Row.names")

design=model.matrix(~0+MEs0$Dominance)
moduleTraitCor = cor(MEs0[,-c(1,2)], design , use = "p");
moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nrow(t(assay(dds_rnaseq_vst))))
textMatrix = paste(signif(moduleTraitCor, 2), "\n(",
                   signif(moduleTraitPvalue, 1), ")", sep = "");
dim(textMatrix) = dim(moduleTraitCor)
par(mar = c(6, 8, 3, 3));
labeledHeatmap(Matrix = moduleTraitCor,
               xLabels = c("Lactobacillus crispatus","Lactobacillus gasseri",
                           "Lactobacillus iners","Others"),
               yLabels = names(MEs0[,-c(1,2)]),
               ySymbols = gsub("ME","",names(MEs0[,-c(1,2)])),
               colorLabels = FALSE,
               colors = blueWhiteRed(50),
               textMatrix = textMatrix,
               setStdMargins = FALSE,
               cex.text = 0.5,
               zlim = c(-1,1),
               main = paste("Module-trait relationships"))
dev.off()

```


Perform EnrichR analysis for each module

```{r}
module_df <- data.frame(
  gene_id = names(netwk$colors),
  colors = labels2colors(netwk$colors)
)
#split by color
module_df_list<-split(module_df$gene_id,module_df$colors)
dbs<-c("KEGG_2021_Human","GO_Biological_Process_2023")

enriched_module_healthy <- lapply(module_df_list, function(x) enrichr(x, dbs))

for (name in names(enriched_module_healthy)) {
  # Construct the file name using the name of the data frame
  file_name <- paste0("data/",name, "_module_EnrichR.txt")
  
  # Write the data frame to a CSV file
  write.table(enriched_module_healthy[[name]]$KEGG_2021_Human, file_name, row.names = FALSE,sep="\t",quote=F )
}
combined_df <- bind_rows(
  lapply(names(enriched_module_healthy), function(category) {
  df <- enriched_module_healthy[[category]]$GO_Biological_Process_2023
  df<- df[1:5,]; df$Category <- category
  df
  })
)

combined_df <- combined_df %>%
mutate(Transformed.P.value = -log10(Adjusted.P.value)) %>% filter(Category!="grey")
combined_df$Category=factor(combined_df$Category, levels = c("yellow","magenta","blue","salmon","greenyellow","green","turquoise","red","pink","tan","brown","purple","black"))
combined_df <- combined_df %>% arrange(Category,Term)
combined_df$Term<-makeUnique(combined_df$Term)
category_colors <- set_names(unique(as.character(combined_df$Category)))

# Create a color palette based on transformed p-values
color_palette <- colorRamp2(range(combined_df$Transformed.P.value), c("blue", "red"))
# Initialize the Circos plot
circos.clear()
circos.par("track.height" = 0.3, start.degree = 90, gap.degree = 0.5)  # Adjust track height and gaps
# Initialize factors
unique_terms <- combined_df$Term
circos.initialize(factors = unique_terms, xlim = c(0, 1))
# Create a track for the category bars
circos.track(
ylim = c(0, 1), bg.border = NA, track.height = 0.05,
panel.fun = function(x, y) {
     sector_index <- get.cell.meta.data("sector.index")
     category_name <- combined_df %>% filter(Term == sector_index) %>% pull(Category)
     # Draw category bar
     circos.rect(
     xleft = 0, ybottom = 0, xright = 1, ytop = 1,
     col = category_colors[category_name], border = NA
          )
     }
)
# Create a barplot track for terms
circos.trackPlotRegion(
     factors = unique_terms, ylim = c(0, max(combined_df$Transformed.P.value)),
     bg.border = NA,
     panel.fun = function(x, y) {
          sector_index <- get.cell.meta.data("sector.index")
          term_data <- combined_df %>% filter(Term == sector_index)
          term_value <- term_data$Transformed.P.value
          category_name <- term_data$Category
          # Use circos.barplot for drawing
          circos.barplot(
               term_value, pos = 0.5,
               col = color_palette(term_value)
          )
          circos.text(
               0.5, term_value + 0.2, term_data$Term_slim,
               facing="reverse.clockwise", niceFacing = TRUE, cex = 0.5
          )
     }
)
lgd_term_value <- Legend(
     at = seq(from = floor(min(combined_df$Transformed.P.value)),
     to = ceiling(max(combined_df$Transformed.P.value)), by = 1),
     col_fun = color_palette,
     title = "Transformed -log10(P-value)"
)
draw(lgd_term_value, x = unit(1, "npc") - unit(30, "mm"), y = unit(1, "mm"), just = c("left", "bottom"))

circos.clear()
#circos.par("track.height" = 0.5, start.degree = 90, gap.degree = 1,canvas.xlim = c(-1.5, 1.5), canvas.ylim = c(-3, 3))  # Adjust track height and gaps
unique_terms <- combined_df$Term
circos.initialize(sectors  = unique_terms, xlim = c(0, 1))
# Create a track for the category bars
circos.track(
     ylim = c(0, 1), bg.border = NA, track.height = 0.05,
     panel.fun = function(x, y) {
          sector_index <- get.cell.meta.data("sector.index")
          category_name <- combined_df %>% filter(Term == sector_index) %>% pull(Category)
          # Draw category bar
          circos.rect(
               xleft = 0, ybottom = 0, xright = 1, ytop = 1,
               col = category_colors[category_name], border = NA
          )
     }
)
# Create a barplot track for terms
circos.trackPlotRegion(
     factors = unique_terms, ylim = c(0, max(combined_df$Transformed.P.value)),
     bg.border = NA,
     panel.fun = function(x, y) {
          sector_index <- get.cell.meta.data("sector.index")
          term_data <- combined_df %>% filter(Term == sector_index)
          term_value <- term_data$Transformed.P.value
          category_name <- term_data$Category
          # Use circos.barplot for drawing
          circos.barplot(
               term_value, pos=0.5,
               col = color_palette(term_value)
          )
     circos.text(
          0.5, term_value + 0.2, term_data$Term_slim,
          facing="reverse.clockwise", niceFacing = TRUE, cex = 0.5
          )
     }
)
lgd_term_value <- Legend(
     at = seq(from = floor(min(combined_df$Transformed.P.value)),
     to = ceiling(max(combined_df$Transformed.P.value)), by = 1),
     col_fun = color_palette,
     title = "Transformed -log10(P-value)"
     )
draw(lgd_term_value, x = unit(1, "npc") - unit(30, "mm"), y = unit(1, "mm"), just = c("left", "bottom"))
circos.clear()
```

Create network for Cytoscape analysis

```{r}
significant_correlations_modules=merge(significant_correlations,module_df,by.x="Gene",by.y="gene_id")
cor_eigen_module=reshape2::melt(WGCNA::cor(MEs0[,-c(1,2)],gene_matrix))
significant_correlations_modules<-merge(significant_correlations_modules,cor_eigen_module,by="Gene")
significant_correlations_modules<-subset(significant_correlations_modules,colors.x==colors.y)
#create network
# Extract unique genes, microorganisms, and colors
unique_genes <- as.character(unique(significant_correlations_modules$Gene))
unique_microorgs <- as.character(unique(significant_correlations_modules$microorg))
unique_colors <- as.character(unique(significant_correlations_modules$colors.x))

# Create a node list data frame
node_list <- data.frame(name = c(unique_genes, unique_microorgs, unique_colors), 
                        shared_name = c(unique_genes, unique_microorgs, unique_colors),
                        Type = c(rep("Gene", length(unique_genes)), 
                                 rep("Microorganism", length(unique_microorgs)), 
                                 rep("Module", length(unique_colors))))

# Create an edge list data frame
edge_list <- data.frame(Source = significant_correlations_modules$Gene, 
                        Target = significant_correlations_modules$microorg, 
                        Correlation = significant_correlations_modules$cor.x, 
                        P_Value = significant_correlations_modules$cor_pval)

# Add edges between genes and modules
gene_module_edges <- data.frame(Source = significant_correlations_modules$Gene, 
                                Target = significant_correlations_modules$colors.x, 
                                Correlation = significant_correlations_modules$cor.y, 
                                P_Value = NA)
edge_list <- rbind(edge_list, gene_module_edges)

# Add edges between microorganisms and modules
microorg_module_edges <- data.frame(Source = significant_correlations_modules$microorg, 
                                    Target = significant_correlations_modules$colors.x, 
                                    Correlation = NA, 
                                    P_Value = NA)
edge_list <- rbind(edge_list, microorg_module_edges)

# Write the node list to a CSV file
write.csv(node_list, file.path(PROJECT_DIR, "results/node_list.csv"), row.names = FALSE,quote = F)

# Write the edge list to a CSV file
write.csv(edge_list, file.path(PROJECT_DIR, "results/edge_list.csv"), row.names = FALSE,quote = F)


```

