---
title: "DataPreparation.Rmd"
author: "Marco Amato Cianci"
date: "2025-04-14"
output: html_document
---

```{r echo=FALSE, warning=FALSE}
suppressMessages(library(DirichletMultinomial))
suppressMessages(library(rstudioapi))
suppressMessages(library(openxlsx))
suppressMessages(library(dplyr))
```

```{r echo=FALSE, warning=FALSE}
script_dir <- dirname(getActiveDocumentContext()$path)
setwd(script_dir)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = script_dir)
```

In this section, we import and preprocess the key dataframes required for the analysis of the **vulvar microbiome**. The datasets include microbiome absolute abundances generated using Kraken, samples metadata, and gene expression counts, all of which are crucial for subsequent statistical and bioinformatic analysis.

-   **Kraken Output**: The dataframe `kdata` represents the taxonomic abundances from the Kraken output. To eliminate potential human DNA contamination, reads associated with *Homo sapiens* have been filtered out.

-   **Metadata**: The `meta` dataframe includes relevant clinical and experimental information about the cohort, containing details for each sample in the study, such as patient/control status.


```{r echo=FALSE}

#Dataframe output Kraken
kdata=read.delim("../local/data/bracken_merged_abbundances.num.txt", sep= "\t")
kdata = kdata[!grepl("sapiens",kdata$name),]
rownames(kdata)=kdata$name
kdata= kdata[,-c(1:3)] 
rownames(kdata)<-gsub("'","_",rownames(kdata))
head(kdata[,1:round(ncol(kdata)/4,0)])

#Dataframe metadata
meta <- read.xlsx("../local/data/Vulvodinia_Samples.xlsx")
kdataf <- sweep(kdata,2,colSums(kdata),"/")
```


### DECONTAMINATION (Decontam Package)
```{r echo=FALSE}
#-----------------------#
# 1. Data Filtering ----
#-----------------------#
# Filter metadata and align with kdata

meta<-meta %>% filter(Analysis.first.draf == "YES")
kdata<-kdata[,unique(paste0(meta$SampleType,"_",meta$PatientID))]

# Prepare decontamination metadata
kdata_dec <- kdata
colnames(kdata_dec)<-sapply(strsplit(colnames(kdata_dec),"_"),"[", 2)

deconmeta<-meta %>% 
  mutate(sample = paste(SampleType,PatientID, sep="_")) %>%
  filter(PatientID %in% colnames(kdata_dec)) %>%         
  dplyr::select(PatientID,sample,`Concentration(ng/ul)-QUBIT`,`Date(libraryready)`,Run,RNA.extraction,`Tape(Library)`,SampleType) %>%
  distinct(PatientID, .keep_all = TRUE) %>%
  filter(`Concentration(ng/ul)-QUBIT` != "NA") %>%
  mutate(`Concentration(ng/ul)-QUBIT` = ifelse(`Concentration(ng/ul)-QUBIT` == "toolow", 0.01, `Concentration(ng/ul)-QUBIT`)) %>%
  tibble::column_to_rownames("sample") %>%
  filter(RNA.extraction %in% names(table(RNA.extraction))[which(table(RNA.extraction) > 1)])


#-----------------------------------------#
# 2. Phyloseq Object Preparation ----
#-----------------------------------------#
## Sample Table ##

samplesdeconmeta<-sample_data(
  data.frame(
    row.names = deconmeta$PatientID,
    SampleID = deconmeta$PatientID,
    Sample_or_Control=c(rep("Sample",nrow(deconmeta))), 
    quant_reading=as.numeric(as.character(deconmeta$`Concentration(ng/ul)-QUBIT`)), 
    run= deconmeta$Run,rnaextr=deconmeta$RNA.extraction, 
    libraries=deconmeta$`Date(libraryready)`,
    SampleType=deconmeta$SampleType
   )
)

## OTU Table ##

kdata_dec <- kdata_dec[,colnames(kdata_dec) %in% rownames(samplesdeconmeta)]
otutable <- otu_table(kdata_dec,taxa_are_rows = TRUE)

# Create Phyloseq object
ps <- phyloseq(otutable, samplesdeconmeta)

#-----------------------------------------#
# 3. Decontamination Analysis ----
#-----------------------------------------#

# Identify contaminants using frequency method
contamdf.freqrnaextr <- isContaminant(
  ps, 
  method="frequency", 
  conc="quant_reading",
  batch="rnaextr",
  threshold=0.1
)

# Filter non-contaminants
rnaextrcontfree <- rownames(filter(contamdf.freqrnaextr, contaminant == FALSE))

#-----------------------------------------#
# 4. Save Clean Data ----
#-----------------------------------------#
x<-sweep(kdata,2,colSums(kdata),"/")
x<-x %>% filter(rownames(x) %in% rnaextrcontfree)
kdata <- kdata %>% 
  filter(rownames(.) %in% rnaextrcontfree)

write.table(x, file = "../results/countsToT_decont_normtss.txt", sep = "\t")
write.table(kdata, file = "../results/countsToT_decont.txt", sep = "\t")
```

### FILTERING 
##### Take the species with more than 0.1% relative abudance in at least 1 sample
```{r echo=FALSE}
##Remove the outliers
kdata <- kdata[,!grepl("BP3",colnames(kdata))]
kdata<-kdata[,!grepl("TT4$",colnames(kdata))]
kdataH<-kdata[,grepl("H_",colnames(kdata))]
total_reads<-colSums(kdataH)
kdataH_frac<-sweep(kdataH, 2, total_reads, "/")
prc<-0.001
idx<- which(rowSums(kdataH_frac >= prc) >= round(ncol(kdataH_frac)/10,0))
kdataH_frac<- as.data.frame(kdataH_frac[idx,])
filtered_micro<-rownames(kdataH_frac)
kdata_filt<-kdataH[filtered_micro,]
write.table(kdata_filt,"../results/counts_decont_filtered.txt", sep="\t")
```