---
title: "PaperMicroUltimate"
output: html_document
date: "2024-10-04"
---

```{r echo=FALSE, warning=FALSE}
suppressMessages(library(DirichletMultinomial))
suppressMessages(library(rstudioapi))
suppressMessages(library(openxlsx))
suppressMessages(library(dplyr))
suppressMessages(library(pheatmap))
suppressMessages(library(ggplot2))
suppressMessages(library(ggrepel))
suppressMessages(library(gridExtra))
suppressMessages(library(patchwork))
suppressMessages(library(reshape2))
suppressMessages(library(colorRamp2))
suppressMessages(library(viridis))
suppressMessages(library(edgeR))
suppressMessages(library(readr))
suppressMessages(library(vegan))
suppressMessages(library(dendextend))
suppressMessages(library(tidyr))
suppressMessages(library(ggExtra))
suppressMessages(library(ggpubr))
suppressMessages(library(phyloseq))
suppressMessages(library(decontam))
suppressMessages(library(waffle))
suppressMessages(library(janitor))
suppressMessages(library(rlang))
suppressMessages(library(svglite))
```

```{r echo=FALSE, warning=FALSE}
script_dir <- dirname(getActiveDocumentContext()$path)
setwd(script_dir)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = script_dir)
```

In this section, we import and preprocess the key dataframes required for the analysis of the **vulvar microbiome**. The datasets include microbiome absolute abundances generated using Kraken, samples metadata, and gene expression counts, all of which are crucial for subsequent statistical and bioinformatic analysis.

-   **Kraken Output**: The dataframe `kdata` represents the taxonomic abundances from the Kraken output. To eliminate potential human DNA contamination, reads associated with *Homo sapiens* have been filtered out.

-   **Metadata**: The `meta` dataframe includes relevant clinical and experimental information about the cohort, containing details for each sample in the study, such as patient/control status.


```{r echo=FALSE}

#Dataframe output Kraken
kdata=read.delim("../local/data/bracken_merged_abbundances.num.txt", sep= "\t")
kdata = kdata[!grepl("sapiens",kdata$name),] # Remove  reads from 'Homo sapiens' 
rownames(kdata)=kdata$name
kdata= kdata[,-c(1:3)] # Remove firsts columns (name,taxonomy_id, taxonomy_lvl)
rownames(kdata)<-gsub("'","_",rownames(kdata))
head(kdata[,1:round(ncol(kdata)/4,0)])

#Dataframe metadata
meta=read.xlsx("../local/data/Vulvodinia_Samples.xlsx") #Metadata sequencing information

#Add H/D information in colnames
# colnames(kdata)=paste(meta[match(colnames(kdata),meta$PatientID),]$SampleType, colnames(kdata), sep="_")
```



### DECONTAMINATION (Decontam Package)
```{r echo=FALSE}

#-----------------------#
# 1. Data Filtering ----
#-----------------------#
# Filter metadata and align with kdata

meta<-meta %>% filter(Analysis.first.draf == "YES")
kdata<-kdata[,unique(paste0(meta$SampleType,"_",meta$PatientID))]

# Prepare decontamination metadata
kdata_dec <- kdata
colnames(kdata_dec)<-sapply(strsplit(colnames(kdata_dec),"_"),"[", 2)

deconmeta<-meta %>% 
  mutate(sample = paste(SampleType,PatientID, sep="_")) %>%
  filter(PatientID %in% colnames(kdata_dec)) %>%         
  dplyr::select(PatientID,sample,`Concentration(ng/ul)-QUBIT`,`Date(libraryready)`,Run,RNA.extraction,`Tape(Library)`,SampleType) %>%
  distinct(PatientID, .keep_all = TRUE) %>%
  filter(`Concentration(ng/ul)-QUBIT` != "NA") %>%
  mutate(`Concentration(ng/ul)-QUBIT` = ifelse(`Concentration(ng/ul)-QUBIT` == "toolow", 0.01, `Concentration(ng/ul)-QUBIT`)) %>%
  tibble::column_to_rownames("sample") %>%
  filter(RNA.extraction %in% names(table(RNA.extraction))[which(table(RNA.extraction) > 1)])


#-----------------------------------------#
# 2. Phyloseq Object Preparation ----
#-----------------------------------------#
## Sample Table ##

samplesdeconmeta<-sample_data(
  data.frame(
    row.names = deconmeta$PatientID,
    SampleID = deconmeta$PatientID,
    Sample_or_Control=c(rep("Sample",nrow(deconmeta))), 
    quant_reading=as.numeric(as.character(deconmeta$`Concentration(ng/ul)-QUBIT`)), 
    run= deconmeta$Run,rnaextr=deconmeta$RNA.extraction, 
    libraries=deconmeta$`Date(libraryready)`,
    SampleType=deconmeta$SampleType
   )
)

## OTU Table ##

kdata_dec <- kdata_dec[,colnames(kdata_dec) %in% rownames(samplesdeconmeta)]
otutable <- otu_table(kdata_dec,taxa_are_rows = TRUE)

# Create Phyloseq object
ps <- phyloseq(otutable, samplesdeconmeta)

#-----------------------------------------#
# 3. Decontamination Analysis ----
#-----------------------------------------#

# Identify contaminants using frequency method
contamdf.freqrnaextr <- isContaminant(
  ps, 
  method="frequency", 
  conc="quant_reading",
  batch="rnaextr",
  threshold=0.1
)

# Filter non-contaminants
rnaextrcontfree <- rownames(filter(contamdf.freqrnaextr, contaminant == FALSE))

#-----------------------------------------#
# 4. Save Clean Data ----
#-----------------------------------------#
kdata <- kdata %>% 
  filter(rownames(.) %in% rnaextrcontfree)

write.table(kdata, file = "../results/decontaminated_counts.txt", sep = "\t")
```
## **Filtering Dataframes Based on Read Counts**

In this step, we apply a filtering criteria to retain only the species with significant read counts in both control (H) and disease (D) groups. Specifically, we select species that have at least 100 reads in 50% or more of the samples within each condition. This filtering process ensures that the analysis focuses on the most abundant and relevant microorganisms across the dataset.

1. **Condition-Specific Filtering**: The columns in the `kdata` dataframe are relabeled based on the sample type (H or D), allowing easy identification of samples belonging to each condition. The dataframe is then split into two subsets: one for disease samples (`kdataD`) and one for healthy controls (`kdataH`).

2. **Selection Criteria**: Species that meet the threshold of 100 reads in at least half of the samples in either the disease or control group are retained. This ensures that only taxa with significant representation across the conditions are included in the analysis.

3. **Final Filtered Dataset**: The filtered dataframe `kdata_filt` contains the selected microorganisms, ordered by their total read counts across all samples, providing a curated list of species for downstream analysis.

This filtering step is essential for focusing the analysis on the most biologically relevant taxa and minimizing noise from low-abundance species.

##### Take the species with more than 100 reads in at least 50% of each condition (H/D)

```{r echo=FALSE}
##Remove the outliers
kdata=kdata[,!grepl("BP3",colnames(kdata))]
kdata=kdata[,!grepl("TT4$",colnames(kdata))]

#Filtering of only healthy samples and only species with a minimum of 100 reads in the half of samples
kdataH<-kdata[,grepl("H_",colnames(kdata))]
#Normalization of dataset based on totnumber of reads (number_of_read/tot_reads_in_sample)
total_reads<-colSums(kdataH)
kdataH_frac<-sweep(kdataH, 2, total_reads, "/")
prc=0.001
idx= which(rowSums(kdataH_frac >= prc) >= round(ncol(kdataH_frac)/10,0))
kdataH_frac= as.data.frame(kdataH_frac[idx,])
filtered_micro<-rownames(kdataH_frac)
kdata_filt<-kdataH[filtered_micro,]

#------------------------------#
# 1. Total Sum Scaling (TSS) ----
#------------------------------#
kdata_frac<-sweep(kdata_filt,2,colSums(kdata_filt),"/")
```


## PCA on the TSS abundances
```{r echo=FALSE}

#Performing a PCA on these data
data_PCA = prcomp(t(kdata_frac), scale = T)
scores = data.frame(sample = rownames(data_PCA$x), PC1 = data_PCA$x[,1], PC2 = data_PCA$x[,2], PC3 = data_PCA$x[,3])
scores$sample = rownames(scores)
scores$status = sapply(strsplit(scores$sample,"_"),"[", 1)
scores$PatientID = sapply(strsplit(scores$sample,"_"),"[", 2)

## **PCA Plot on PC1 vs PC2**
var_explained <- (data_PCA$sdev^2) / sum(data_PCA$sdev^2)
pca_plot <- ggplot(scores, aes(x = PC1, y = PC2 )) +
  geom_point(size = 3, alpha = 1,color="#7EB77F") +  # Points with size and transparency
  theme_minimal(base_size = 14) +     # Minimal theme for a clean design
    # Add text labels with smart positioning
  ggrepel::geom_text_repel(
    aes(label = sample),  # Replace 'SampleID' with your actual column name
    color = "#2D5C2F",       # Darker green for contrast
    size = 2,
    box.padding = 0.35,      # Prevent labels from overlapping points
    point.padding = 0.3,
    max.overlaps = 20,       # Adjust based on your data density
    segment.color = "gray60",
    min.segment.length = 0.2
  ) +
  labs(
    title = "PCA Plot of Samples",
    x = paste0("PC1 (", round(var_explained[1] * 100, 2), "% Variance)"),
    y = paste0("PC2 (", round(var_explained[2] * 100, 2), "% Variance)"),
    color = "Status",
    shape = "Patient ID"
  ) +
  theme_minimal()+
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16), # Centered title
    axis.text = element_text(size = 12),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10)
  )
pca_plot
```

## NMDS on TSS data
```{r echo=FALSE, message=FALSE, warning=FALSE}

NMDS <- vegan::metaMDS(
  t(kdata_frac), 
  distance = "bray", 
  trymax = 100,
  autotransform = TRUE,  # Auto-transformation enabled (Wisconsin double standardization)
  k = 4                  # Number of dimensions
)


# Extract and prepare scores
stress_value <- paste("Stress:", round(NMDS$stress, 3))  # Capture stress value
score <- data.frame(
  vegan::scores(NMDS)$site,
  sampleID = rownames(vegan::scores(NMDS)$site)
) %>%
  mutate(
    SampleGroup = sapply(strsplit(sampleID, "\\."), "[", 1)  # Extract sample groups
  )

ggplot(score, aes(x = NMDS1, y = NMDS2)) +
  geom_point(
    size = 3,
    alpha = 0.9,
    color = "#7EB77F",      # Base color
    fill = "#7EB77F30",     # Semi-transparent fill
    shape = 21,             # Filled circle shape
    stroke = 0.7            # Border thickness
  ) +
  ggrepel::geom_text_repel(
    aes(label = SampleGroup),
    color = "#2D5C2F",
    size = 2,
    box.padding = 0.4,
    point.padding = 0.3,
    max.overlaps = 15,
    segment.color = "grey40",
    min.segment.length = 0.15,
    seed = 123              # For reproducible label positions
  ) +
  labs(
    title = "Non-Metric Multidimensional Scaling (NMDS)",
    subtitle = stress_value,
    x = "NMDS Axis 1",
    y = "NMDS Axis 2"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    axis.title = element_text(face = "bold", color = "gray30"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  coord_fixed()  # Maintain aspect ratio

```


# Clustering 

```{r echo=FALSE}
### Matrice di Distanze ###

beta_rna <- vegdist(t(kdata_frac), method = "bray")
# beta_rna <- dist(t(b))
betamtx<-as.matrix(beta_rna)
pheatmap(betamtx, 
         # color = viridis,
         clustering_distance_rows = beta_rna,
         clustering_distance_cols = beta_rna)


###HIERARCHICAL CLUSTERING UNSCALED###
library(phylogram)

hc <- hclust(dist(t(kdata_frac),method = "euclidean"), method = "ward.D2")
clusters <- cutree(hc, k = 4)
### Dendrogramma ###
dend <- as.dendrogram(hc)

colored_dend <- color_branches(dend, k = 4, col = c("#F58B1F", "#92C73E","#EC1E24", "#468B74"  , "#35459B"))
plot(colored_dend, type = "triangle", ylab = "Height")
write.dendrogram(dend, file = "../results/Dendrogram.nwk", edges = TRUE)

#Labels for later orders
ordered_samples<-labels(dend)
ordereds_samples_2=sapply(strsplit(ordered_samples,"_"),"[",2)
kdata_frac=kdata_frac[rowSums(kdata_frac)!=0,match(ordered_samples,colnames(kdata_frac))]


### Setting Dominance of CST found in Vaginal microbiome
species=c("Lactobacillus iners","Lactobacillus crispatus","Lactobacillus gasseri","Lactobacillus jensenii")
custom_colors <- c(
   "Lactobacillus crispatus" = "#F58B1F",
  "Lactobacillus iners" = "#92C73E",  
  "Lactobacillus gasseri" = "#EC1E24",
  "Others" = "#468B74",  
  "Lactobacillus jensenii" = "#35459B"
  )

### Creating Dominance Dataframe
lacto=as.data.frame(kdata_frac[rownames(kdata_frac) %in% species,])
cstdata=rbind(lacto,Others=1-colSums(lacto))
cstdata$Species <- rownames(cstdata)

data_long <- cstdata %>%
  pivot_longer(
    cols = starts_with("H_"),   # Specify the columns to pivot
    names_to = "Category",       # Name of the new column for the original column names
    values_to = "Value"          # Name of the new column for the values
  ) 
data_long$Category<-factor(data_long$Category,levels = ordered_samples)


#### Plot Abbondanze ####

ggplot(data_long, aes(y = Value, x = Category, fill = Species,)) +
  geom_col(width = 0.98) +
   scale_fill_manual(values = custom_colors) + 
  #geom_text(vjust = -0.5) +  # Aggiungi le etichette sopra le barre
  theme_minimal()  # Usa un tema semplice


#### Heatmap ####

annotationcol=data_long %>% group_by(Category) %>% filter(Value==max(Value)) 
annotationcol=data.frame(row.names = annotationcol$Category, Dominance=annotationcol$Species)

cluster_col=data.frame(row.names = names(clusters),Clusters=clusters)#,Clusters_scaled=clusters_scaled)
cluster_col$Clusters=paste0("cl_",cluster_col$Clusters)
# cluster_col$Clusters_scaled=paste0("cl_",cluster_col$Clusters_scaled)

final_annotationcol=merge(annotationcol,cluster_col,by=0)
rownames(final_annotationcol)=final_annotationcol$Row.names
final_annotationcol$Row.names=NULL
final_annotationcol<-arrange(final_annotationcol,Clusters)

#Clusters_scaled =  c("cl_1"="#264653",  "cl_2"="#2A9D8F","cl_3"="#E9C46A", "cl_4"="#F4A261" )

cluster_colors <- list(Clusters = c("cl_1" = "#468B74" , "cl_2" ="#F58B1F", "cl_3" = "#92C73E","cl_4" =  "#EC1E24"), Dominance=  c("Lactobacillus crispatus" = "#F58B1F", "Lactobacillus gasseri" = "#EC1E24","Lactobacillus iners" = "#92C73E",  "Others" = "#468B74"  ,   "Lactobacillus jensenii" = "#35459B"))



pheatmap(kdata_frac, 
         cluster_rows = T, 
         cluster_cols = F,
         annotation_colors  = cluster_colors,
         annotation_col = final_annotationcol,
         scale= "row",
         fontsize_row = 8, 
         fontsize_col = 8,
         clustering_method = "ward.D2",
         angle_col = 45,
         color = viridis(100),
         breaks= seq(-1.2,1.5,length.out=100)
         
)



```

```{r Dams_cluster}

mann_whitney <- function(taxa,group) {
  group1<-rownames(subset(cluster_col,Clusters==group))
  group2<-rownames(subset(cluster_col,Clusters!=group))
  test <- wilcox.test(taxa[group1], taxa[group2])
  median_change <- median(taxa[group1]) - median(taxa[group2])
  return(data.frame(pvalue=test$p.value,logMedianChange=median_change))
}

dgeobj <- DGEList(counts=kdata_filt)
dgeobj <- calcNormFactors(dgeobj,method="TMM")
counts_norm <- cpm(dgeobj, normalized.lib.sizes=TRUE, log=TRUE)


DGE<-lapply(unique(cluster_col$Clusters), function(cluster) { 
           results<-apply(counts_norm, 1, function(taxa) mann_whitney(taxa,cluster))
           results<- do.call(rbind,results)
           results$padj<-p.adjust(results$pvalue, method = "BH") 
           return(results)
             })

names(DGE)=unique(cluster_col$Clusters)
DGE_filt<- lapply(DGE,function(x) rownames(subset(x,padj<=0.1 & abs(logMedianChange)>=0.5)))
DGE_filt_list <- unique(do.call(c,DGE_filt)) 

pheatmap(counts_norm[DGE_filt_list,match(ordered_samples,colnames(counts_norm))], 
         cluster_rows = T, 
         cluster_cols = F,
         annotation_colors  = cluster_colors,
         annotation_col = final_annotationcol,
         scale= "row",
         fontsize_row = 10, 
         fontsize_col = 10,
         clustering_method = "ward.D2",
         # cutree_cols = 5,
         cutree_rows = 4,
         angle_col = 45,
         color = viridis(100),
         breaks= seq(-1.2,1.5,length.out=100)
)
```


```{r}

bac2<-rowSums(apply(kdata_frac,2,FUN = function(x) x>=0.1))>=1
genus_frac_filt2<-kdata_frac[bac2,]
genus_frac_filt2["<10%",]<-1-colSums(genus_frac_filt2)
mypalette=c("grey80","#f65356","#fb8066","#fea671","#fec979","#feea80","#d7e586","#a5de94","#5daad8","#8187c7","#c56bba")

inerssamples<-rownames(annotationcol)[which(annotationcol$Dominance == "Lactobacillus iners")]
inersfrac<-genus_frac_filt2[,match(inerssamples,colnames(genus_frac_filt2))]
inersfrac$means<-rowMeans(inersfrac)
crispatussamples<-rownames(annotationcol)[which(annotationcol$Dominance == "Lactobacillus crispatus")]
crispatusfrac<-genus_frac_filt2[,match(crispatussamples,colnames(genus_frac_filt2))]
crispatusfrac$means<-rowMeans(crispatusfrac)
gasserisamples<-rownames(annotationcol)[which(annotationcol$Dominance == "Lactobacillus gasseri")]
gasserifrac<-genus_frac_filt2[,match(gasserisamples,colnames(genus_frac_filt2))]
gasserifrac$means<-rowMeans(gasserifrac)
otherssamples<-rownames(annotationcol)[which(annotationcol$Dominance == "Others")]
othersfrac<-genus_frac_filt2[,match(otherssamples,colnames(genus_frac_filt2))]
othersfrac$means<-rowMeans(othersfrac)

inplot<-ggplot(data=inersfrac,aes(x="",y=means,fill = rownames(inersfrac)))+
  geom_bar(stat="identity",width = 1)+
  scale_fill_manual(values = mypalette)+
  coord_polar("y",start=0)+
  xlab(label = "")+
  labs(fill="Species")+
  theme_minimal()

crisplot<-ggplot(data=crispatusfrac,aes(x="",y=means,fill = rownames(crispatusfrac)))+
  geom_bar(stat="identity",width = 1)+
  scale_fill_manual(values = mypalette)+
  coord_polar("y",start=0)+
    xlab(label = "")+
    labs(fill="Species")+
  theme_minimal()

gassplot<-ggplot(data=gasserifrac,aes(x="",y=means,fill = rownames(gasserifrac)))+
  geom_bar(stat="identity",width = 1)+
  scale_fill_manual(values = mypalette)+
  coord_polar("y",start=0)+
    labs(fill="Species")+
    xlab(label = "")+
  theme_minimal()


othersplot<-ggplot(data=othersfrac,aes(x="",y=means,fill = rownames(inersfrac)))+
  geom_bar(stat="identity",width = 1)+
  scale_fill_manual(values = mypalette)+
  coord_polar("y",start=0)+
    labs(fill="Species")+
    xlab(label = "")+
  theme_minimal()

ggarrange(crisplot,inplot, gassplot, othersplot,
                    labels = c("CSTI", "CSTIII", "CSTII","CSTIV"),
                    ncol = 2, nrow = 2)
```

```{r}
genus_frac_filt2$Species <- rownames(genus_frac_filt2)

#### Plot Abbondanze ####
data_long_10prc <- genus_frac_filt2 %>%
  pivot_longer(
    cols = starts_with("H_"),   # Specify the columns to pivot
    names_to = "Category",       # Name of the new column for the original column names
    values_to = "Value"          # Name of the new column for the values
  ) 

library(forcats)

# # Assicurati che "Others" sia l'ultimo livello del fattore Species
#     data_long_10prc$Species <- fct_relevel(data_long_10prc$Species, c("Others"), after = 0)
#     data_long_10prc$Species <- fct_relevel(data_long_10prc$Species, c("Lactobacillus crispatus"), after = 1)
#     data_long_10prc$Species <- fct_relevel(data_long_10prc$Species, c("Lactobacillus iners"), after = 2)
#     data_long_10prc$Species <- fct_relevel(data_long_10prc$Species, c("Lactobacillus jensenii"), after = 3)
# 
data_long_10prc$Category<-factor(data_long_10prc$Category,levels = ordered_samples)
ggplot(data_long_10prc[,], aes(y = Value, x = Category, fill = Species)) +
  geom_col() +
  scale_fill_manual(values = mypalette)+
  theme_minimal()  # Usa un tema semplice

```
```{r}
data_PCA = prcomp(t(kdata_frac), scale = T)
scores = data.frame(sample = rownames(data_PCA$x), PC1 = data_PCA$x[,1], PC2 = data_PCA$x[,2], PC3 = data_PCA$x[,3])
scores$sample = rownames(scores)
scores$status = sapply(strsplit(scores$sample,"_"),"[", 1)
scores$PatientID = sapply(strsplit(scores$sample,"_"),"[", 2)
scores = merge(scores,final_annotationcol,by = 0)
```

---

### **PCA Plot on PC1 vs PC2**

```{r echo=FALSE, warning=FALSE, message=FALSE}
var_explained <- (data_PCA$sdev^2) / sum(data_PCA$sdev^2)
p=ggplot(scores, aes(x=PC1, y=PC2,color= Dominance, label= PatientID)) +
 # ggtitle("PCA su filtro per 100reads in 50% dei campioni (x condizione)")+
  geom_point(size = 3) +
  scale_color_manual(values = cluster_colors$Dominance) +
  geom_text_repel(max.overlaps = 10, size= 3, box.padding = 0.3, point.padding = 3) +
   # Minimal theme for a clean design
  labs(
    title = "PCA Plot of Samples",
    x = paste0("PC1 (", round(var_explained[1] * 100, 2), "% Variance)"),
    y = paste0("PC2 (", round(var_explained[2] * 100, 2), "% Variance)"),
    color = "Status",
    shape = "Patient ID"
  ) +
  theme_test()
pmarginal <- ggMarginal(p, type = "density", groupColour = TRUE, groupFill = TRUE)
  

 ggsave(filename = paste0("../results/images/PCAClusters.svg"), plot = p, width = 8, height = 6, device = svglite)
```

```{r alphadiversity}
nreads_min<-min(colSums(kdata_filt))
nreads_max<-max(colSums(kdata_filt))
fold<-round(nreads_max/nreads_min,0)

kdata_subs<-rrarefy(t(kdata_filt),nreads_min)
alpha_subs<-diversity(kdata_subs)#t(kdata[,match(ordered_samples,colnames(kdata))])
alphasubs=data.frame(samples=names(alpha_subs),value=alpha_subs)
alphasubs$samples<-factor(alphasubs$samples,ordered_samples)
alphasubs<-merge(alphasubs,final_annotationcol,by=0) %>% tibble::column_to_rownames("Row.names")

my_comparisons <- list( c("cl_1","cl_2"),c("cl_1","cl_3"),c("cl_1","cl_4"),c("cl_3","cl_2"),c("cl_4","cl_2"),c("cl_3","cl_4"))

ggplot(alphasubs,aes(x=Clusters,y=value,fill = Clusters))+
  geom_boxplot()+  
  scale_fill_manual(values = c("cl_1" = "#468B74" , "cl_2" ="#F58B1F", "cl_3" = "#92C73E","cl_4" =  "#EC1E24"))+
  stat_compare_means(comparisons=my_comparisons)+
  theme_bw()

ggplot(alphasubs, aes(x = samples, y = factor(1), size = value, colour = value)) +
  geom_point() +
  scale_size_continuous(range = c(5, 15)) +
  scale_color_gradient(low = "grey90", high = "#D94943") +  # Intensity of color
  theme_bw() +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none"  # Hide the legend
  )+
  coord_fixed(ratio = 2)

```



```{r}
genus_frac<-kdata_frac
genus_frac$genus<-sapply(strsplit(rownames(genus_frac)," "),"[",1) 
genus_frac<-genus_frac %>%
  group_by(genus) %>%
  summarise(across(everything(), sum, na.rm = TRUE)
  )%>% tibble::column_to_rownames("genus")

genus_frac$sum <- rowSums(genus_frac)

genus_frac<-arrange(genus_frac,desc(sum))
genus_frac$sum<-NULL

prc=0.03
#--- Show only the ones over 20%
bac<-rowSums(apply(genus_frac,2,FUN = function(x) x>= prc))>=1
genus_frac_filt<-genus_frac[bac,]


correlation<-cor(t(genus_frac_filt))
pheatmap(correlation)

```

```{r}

mypalette<-c("#aaa","#dc89ba","#94479c","#54478c" ,"#2c699a", "#048ba8", "#0db39e", "#83e377", "#b9e769", "#efea5a", "#f1c453", "#f29e4c","#D94943","#a4161a","#61161a","#21161a")
genus_frac_filt$Genus <- rownames(genus_frac_filt)

#### Plot Abbondanze ####
data_long_gen <- genus_frac_filt %>%
  pivot_longer(
    cols = starts_with("H_"),   # Specify the columns to pivot
    names_to = "Category",       # Name of the new column for the original column names
    values_to = "Value"          # Name of the new column for the values
  ) 


data_long_gen<-merge(data_long_gen,final_annotationcol,by.x="Category",by.y=0)
data_long_gen$Category<-factor(data_long_gen$Category,levels = ordered_samples)

###ORDINA PER CLUSTER
ggplot(data_long_gen, aes(y = Value, x = Category, fill = Genus)) +
  geom_col(width = 0.98) +
  scale_fill_manual(values = mypalette)+
  theme_minimal()  # Usa un tema semplice

data_long_gen<-subset(data_long_gen,Genus!="Lactobacillus")

ggplot(data_long_gen, aes(x = Genus, y = Category, size = Value, fill = Genus)) +
  geom_point(alpha = 0.7, shape = 21) +
  ggtitle(label=paste0("BubblePlot ",prc*100,"%"))+
  scale_fill_manual(values = mypalette)+
  scale_size_continuous(range = c(0, 27)) +  # Adjust the range for bubble sizes
  theme_minimal()+ 
  theme(legend.position = "right",axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))  # Position the legend as needed

colSums(genus_frac_filt[,-ncol(genus_frac_filt)])
# Pairwise comparisons: Specify the comparisons you want
my_comparisons <- list( c("Lactobacillus crispatus", "Lactobacillus iners"), c("Lactobacillus iners", "Lactobacillus gasseri"), c("Lactobacillus crispatus", "Others"),c("Lactobacillus gasseri", "Others"),c("Lactobacillus iners", "Others"),c("Lactobacillus crispatus", "Lactobacillus gasseri"))


data_long_gen$Cluster<-final_annotationcol[match(data_long_gen$Category,rownames(final_annotationcol)),]$Dominance

ggplot(data_long_gen,aes(x = Cluster,y=Value,fill = Cluster))+
  geom_boxplot()+
  geom_jitter(alpha=0.9)+
  facet_wrap(~Genus,scales="free")+
  stat_compare_means()+
  theme_minimal()

```
```{r}

association <- read.csv("../local/data/MicroPhenoDBAssociationScores.csv") %>% select(-Pmid, -Association_ID)
key_cols <- setdiff(names(association), "Score")


merged_dupl <- merge(kdata_frac, association, by.x = 0, by.y = "Microbe")

# Compute the mean of 'Score' while grouping by 'Row.names'
Scores <- merged_dupl %>%
    group_by(Row.names) %>%
    summarise(Score = mean(Score, na.rm = TRUE), .groups = "drop") %>%
    tibble::column_to_rownames("Row.names")

merged <- merge(kdata_frac,Scores,by=0)%>%tibble::column_to_rownames("Row.names")


diseasescore<-list()
for (names in setdiff(colnames(merged), c("Score","Genus"))) {

diseasescore[names]<-sum(merged[,names]*merged[,"Score"])  
  
}

disbiodb=data.frame(samples=names(unlist(diseasescore)),dysbio=unlist(diseasescore))
disbiodb$samples=factor(disbiodb$samples,ordered_samples)
disbiodb$dysbio<-disbiodb$dysbio

ggplot(disbiodb, aes(x = samples, y = factor(1), size = dysbio, colour = dysbio)) +
  geom_point() +
  scale_size_continuous(range = c(5, 15)) +
  scale_color_gradient(low = "grey90", high = "#0070BB") +  # Intensity of color
  theme_bw() +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    legend.position = "none"  # Hide the legend
  )+
  coord_fixed(ratio = 2)

write.csv(merge(final_annotationcol,disbiodb,by=0) %>% tibble::column_to_rownames("Row.names"),"../results/cluster_col.csv")

```
```{r}
library(ggthemes)
x <- merge(disbiodb,alphasubs)


# Calculate the correlation coefficient
correlation <- cor(x$value, x$dysbio)

# Create the plot
plot <- ggplot(x, aes(x = value, y = dysbio)) +
  # Add points with a refined aesthetic
  geom_point(size = 3.5, alpha = 0.9, shape = 21, fill = "#34495E") +
  
  # Add a smooth regression line with confidence interval
  geom_smooth(method = "lm", se = TRUE, color = "#E74C3C", fill = "#FADBD8", size = 1.5) +
  
  # Add dynamic annotation for correlation score
  annotate("text", x = max(x$value) * 0.8, y = max(x$dysbio) * 0.9,
           label = paste("Correlation: r =", round(correlation, 2)),
           color = "#34495E", size = 5, fontface = "italic") +
  
  # Add custom labels with refined typography
  labs(
    title = "Correlation AlphaDiversity and Dysbiosis",
    subtitle = "scatterplot and regression line",
    x = "Alpha Diversity",
    y = "Dysbiosis Score"
  ) +
  
  # Apply a sophisticated theme for professional presentation
  theme_bw(base_size = 16) +
  
  # Customize theme elements for visual hierarchy
  theme(
    panel.grid.major = element_line(color = "grey85"),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(color = "#34495E", size = 1),
    axis.title.x = element_text(size = 14, face = "bold", margin = margin(t = 10)),
    axis.title.y = element_text(size = 14, face = "bold", margin = margin(r = 10)),
    axis.text.x = element_text(size=12),
    axis.text.y=element_text(size=12),
    plot.title=element_text(size=20, face="bold", hjust=0.5, color="#2C3E50"),
    plot.subtitle=element_text(size=16, hjust=0.5, color="#7F8C8D"),
    legend.position="none"
  )


 ggsave(filename = "../results/images/corralphadisbyo.svg", plot = plot, width = 8, height = 6,device = svglite)
```